---
title: "Effects of Severe Weather by Event Type"
author: "Pat Mitoraj"
date: "October 16, 2017"
output: html_document
---
##Synopsis
This analysis look at storm data from 1996 through November 2011 in order to find which event types were most harmful in terms of population health and economic damage. This analysis is based on the the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, which tracks characteristics of major storms and weather events in the United States. Based on this data, tornadoes are found to be the most harmful to population health, while rain related weather events cause the greatest amount of economic damage.

##Data Processing
The data is first downloaded from the NOAA website, decompressed, and loaded into R. Additionally, all necessary packages for the analysis are installed and loaded:
```{r}
library(R.utils)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
```
```{r, cache=TRUE}
datafile<-if(!file.exists("./repdata%2Fdata%2FStormData.csv")){bunzip2("repdata%2Fdata%2FStormData.csv.bz2",remove=FALSE)}
df<-read.csv("repdata%2Fdata%2FStormData.csv")
```
The dataset tracks data beginning in 1950, however all event types were not recorded until January 1, 1996. In order to accurately compare the damage caused by the different event types, all the events before 1996 are dropped. This is done by first separating BGN_DATE column so that the beginning date of each storm is isolated. Then, I find which row contains the event on 1/1/1996, and drop all preceding rows from the dataset.
```{r}
newdf<-df %>% separate(BGN_DATE, c("BGN_DATE", "DROP"), " ")
newdf$DROP<-NULL
which(grepl("1/1/1996", newdf$BGN_DATE))
newdf2<-newdf[-c(1:249454),]
```
Next, all the dataset features not needed for this analysis are removed.
``` {r}
delete<-c("STATE__", "BGN_TIME", "TIME_ZONE","COUNTY","COUNTYNAME","STATE","END_DATE","END_TIME","COUNTY_END","COUNTYENDN","WFO","STATEOFFIC","ZONENAMES","LATITUDE","LONGITUDE","LATITUDE_E","LONGITUDE_","REMARKS")
newdf2<- newdf2[, !(names(newdf2) %in% delete)] 
```
The property damage and crop damage columns then had to be transformed for use in the analysis. The PROPDMGEXP and CROPDMGEXP columns contained information on the units of damage, but these were coded as exponent values. These columns were mapped to their corresponding numeric values and multiplied by the PROPDMG and CROPDMG columns to get the respective damage amounts as numeric values.
```{r}
newdf2$PROPDMGEXP<- mapvalues(newdf2$PROPDMGEXP, from = c("K", "M","", "B", "m", "+", "0", "5", "6", "?", "4", "2", "3", "h", "7", "H", "-", "1", "8"), to = c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6, 0, 10^4, 10^2, 10^3, 10^2, 10^7, 10^2, 0, 10, 10^8))
newdf2$PROPDMGEXP<- as.numeric(as.character(newdf2$PROPDMGEXP))
newdf2$PROPDMGTOTAL<- (newdf2$PROPDMG * newdf2$PROPDMGEXP)/1000000000
newdf2$CROPDMGEXP<- mapvalues(newdf2$CROPDMGEXP, from = c("","M", "K", "m", "B", "?", "0", "k","2"), to = c(1,10^6, 10^3, 10^6, 10^9, 0, 1, 10^3, 10^2))
newdf2$CROPDMGEXP<- as.numeric(as.character(newdf2$CROPDMGEXP))
newdf2$CROPDMGTOTAL<- (newdf2$CROPDMG * newdf2$CROPDMGEXP)/1000000000
newdf2$DAMAGETOTAL<- newdf2$PROPDMGTOTAL + newdf2$CROPDMGEXP
```
The final step in cleaning the data was grouping the weather events by event type. Many of the events were of the same type but coded differently. In order to fix this problem, events of similar type were grouped together.
```{r}
storm_type<- newdf2 %>%
  mutate(evtypegrp = ifelse(grepl("LIGHTNING|LIGNTNING", EVTYPE), "LIGHTNING", ifelse(grepl("HAIL", EVTYPE), "HAIL", ifelse(grepl("RAIN|FLOOD|WET|FLD", EVTYPE), "RAIN", ifelse(grepl("SNOW|WINTER|WINTRY|BLIZZARD|SLEET|COLD|ICE|FREEZE|AVALANCHE|ICY", EVTYPE), "WINTER",
                                                                                                                                                                      ifelse(grepl("TORNADO|FUNNEL", EVTYPE), "TORNADO",
                                                                                                                                                                                       ifelse(grepl("WIND|HURRICANE", EVTYPE), "WINDS",
                                                                                                                                                                                              ifelse(grepl("STORM|THUNDER|TSTM|TROPICAL +STORM", EVTYPE), "STORM",
                                                                                                                                                                                                     ifelse(grepl("FIRE", EVTYPE), "FIRE",
                                                                                                                                                                                                            ifelse(grepl("FOG|VISIBILITY|DARK|DUST", EVTYPE), "FOG",
                                                                                                                                                                                                                   ifelse(grepl("WAVE|SURF|SURGE|TIDE|TSUNAMI|CURRENT|SWELL", EVTYPE), "WAVE",
                                                                                                                                                                                                                          ifelse(grepl("HEAT|HIGH +TEMP|RECORD +TEMP|WARM|DRY", EVTYPE), "HEAT",
                                                                                                                                                                                                                                 ifelse(grepl("VOLCAN", EVTYPE), "VOLCANO",
                                                                                                                                                                                                                                        ifelse(grepl("DROUGHT", EVTYPE), "DROUGHT","OTHER"))))))))))))))
```
## Results
First the results are summarized in a table showing the total damage, population health dameg, crop damage, injuries, and fatalities by event type.
```{r}
event_sum<- storm_type %>% group_by(evtypegrp) %>% summarise(damage = sum(DAMAGETOTAL), property= sum(PROPDMGTOTAL), crops = sum(CROPDMGTOTAL), fatalities = sum(FATALITIES), injuries = sum(INJURIES))
event_sum
```

Graphing the number of fatalities by event type, we can see that heat related events caused the most fatalities by a significant amount.
```{r}
fatalities_graph <-ggplot(event_sum, aes(evtypegrp, fatalities))+
  geom_bar(stat="identity")+
  ylab("Number of Fatalities")+
  xlab("Event Type")+
  ggtitle("Fatalities by Storm Type (1996-November 2011)")+
  geom_text(aes(label=fatalities), position=position_dodge(width=0.9), vjust=-0.25)
fatalities_graph
```

However, when looking at injuries, tornadoes were the largest cause of damage.
```{r}
injuries_graph <-ggplot(event_sum, aes(evtypegrp, injuries))+
  geom_bar(stat="identity")+
  ylab("Number of Injuries")+
  xlab("Event Type")+
  ggtitle("Injuries by Storm Type (1996-November 2011)")+
  geom_text(aes(label=injuries), position=position_dodge(width=0.9), vjust=-0.25)
injuries_graph
```

Taking into consideration both fatalities and injuries, tornadoes were most harmful to population health. In terms of economic damage, the following figure shows both the property damage and crop damage caused by each event type.
```{r}
damage <-head(event_sum[order(event_sum$damage, decreasing=TRUE),],5)
property <- damage %>% mutate(damage_type="Property", damage_amount=property)
crops <- damage %>% mutate(damage_type="Crops", damage_amount=crops)
damage_major <- rbind(property,crops)
ggplot(damage_major, aes(evtypegrp, damage_amount, fill=factor(damage_type))) +
  geom_bar(stat = "identity") + 
  ylab("Economic damage 1996 to November 2011") +
  xlab("Event Type") +
  scale_fill_discrete(name = "Damage") +
  ggtitle ("Total Economic Damage by Storm Type") +
  theme_gray(base_size = 10, base_family = "")
```

These results indicate that rain related events were the most harmful event in terms of economic damage.

##Conclusion
In summary, this analysis finds that, form 1996- November 2011, tornadoes were the most harmful storm type in terms of population health, while rain events caused the most economic damage.